---
title: "Quantium Virtual Intership Task 1"
author: "Lizbeth Santiago"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Background information

The Category Manager for Chips wants understand the types of customers who purchase Chips and their purchasing behaviour within the region.
The insights from the analysis will feed into the supermarket’s strategic plan for the chip category in the next half year.

## Exploratory data analysis

### Load required libraries

```{r, warning=FALSE}
library(data.table)
library(ggplot2)
library(readr)
library(ggmosaic)
```

### Reading data

```{r}
filePath <- "C:/Users/Liz/Documents/Virtual_Intership/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
```


### Examining transaction data


```{r}
head(transactionData)
str(transactionData)
```

The column DATE is in an integer format, so it is necessary to change this to a date format

### Convert column DATE to date format

```{r}
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

### Verify information - examine PROD_NAME


To check if the products are those that the analysis need.
Choose the unique product names, and examine the words in PROD_NAME to see if there are any incorrect entries such as products that are not chips

```{r}
#unique(transactionData[, PROD_NAME])

productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), " ")))
setnames(productWords, 'words') #name the column
```

### Removing digits and special character


To focus just in words that will tell us if the product is chips or not, remove all words with digits and special characters.

```{r}
patterns <- c('^[123456789]','&')
words2 <- productWords [! grepl (paste(patterns, collapse='|'), productWords$words),]
```

### Most common words and sorting them

```{r}
commonWords <- words2[, .(.N), .(words)][order(-N)]
```

There are some salsa products in the dataset that we don't need.

### Remove Salsa products

```{r}
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]# new column that marks a salsa product
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

### Summarise the data to check for nulls and possible outliers

```{r}
summary(transactionData) #there is a case where 200 packets of chips are bought in one transaction.
```

There is a case where 200 packets of chips are bought in one transaction.

```{r}
transactionData[PROD_QTY>10]
transactionData[LYLTY_CARD_NBR==226000]
```

Actually we found that there are two transactions by the same customer, with card number 226000, who just made these two transactions of 200 packets.

### Removing these two transactions

```{r}
transactionData <- transactionData[LYLTY_CARD_NBR!=226000]
```

### Re-examine transaction data

```{r}
summary(transactionData)
```

### Count the number of transactions by date

```{r}
countByDate<- transactionData[, .(.N), .(DATE)]#a missing date
```

There is a missing date, so create a sequence of dates and use this to create a chart of number of transactions over time to find the missing date.

```{r}
s <- as.Date('2018-07-01')
e <- as.Date('2019-06-30')
completeDates <- data.table(seq(from=s, to=e, by=1))

countByDate2 <- merge(countByDate, completeDates,by.x="DATE", by.y="V1", all=T)
```

### Plot transactions over time


```{r}
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(countByDate2, aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


We can see that there is an increase in purchases in December and a break in late December. Let's zoom in on this.

### Ploting transaction of December and January

```{r}
dec_jan = countByDate2[DATE>as.Date('2018-12-07') & DATE<as.Date('2019-01-07')]

ggplot(dec_jan, aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

The missing date is on December 25th, that day shops are closed.
We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself, then decrease to the usual sales.

## Checking information about Pack size

```{r}
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]
pack_sizes <- transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]

ggplot(transactionData, aes(x = PACK_SIZE)) +
  geom_histogram(binwidth = 10, col='black', fill='green', alpha=0.4) +
  labs(x = "Pack Size", y = "Number of transactions", title = "Transactions by pack size")
```

Most of the sales are from 175g pack size

## Checking Brands

To obtain the brands we extract just the first word of product names

```{r, warning=FALSE}
library(stringr)
transactionData[, BRAND := str_split(transactionData$PROD_NAME, " ", simplify = TRUE)[,1] ] #First word
unique(transactionData[, BRAND])
```

Some of the brands have wrong spelling

### Brand adjustments

```{r}
transactionData[BRAND == "Red", BRAND := "RRD"]
transactionData[BRAND == "Smith", BRAND := "Smiths"]
transactionData[BRAND == "Dorito", BRAND := "Doritos"]
transactionData[BRAND == "Infzns", BRAND := "Infuzions"]
transactionData[BRAND == "Snbts", BRAND := "Sunbites"]
transactionData[BRAND == "WW", BRAND := "Woolworths"]
transactionData[BRAND == "NCC", BRAND := "Natural"]
transactionData[BRAND == "Grain", BRAND := "GrnWves"]
```

### Re-examine of brands

```{r}
unique(transactionData[, BRAND])
```

## Examining customer data

```{r}
head(customerData)
summary(customerData)
```

## Merge both datasets

```{r}
data <- merge(transactionData, customerData, all.x = TRUE)
```

### Checking for nulls.

```{r}
sum(is.na(data))
```

### Save 'data'

```{r}
#filePath <- "C:/Users/Liz/Documents/Virtual_Intership/"
#fwrite(data, paste0(filePath,"QVI_data.csv"))
```
