---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 2"
author: "Lizbeth Santiago"
output:
  pdf_document: default
  html_document: default
---

The client has selected store numbers 77, 86 and 88 as trial stores and want control stores to be established stores that are operational for the entire observation period.
We would want to match trial stores to control stores that are similar to the trial store prior to the trial period of Feb 2019 in terms of :

* Monthly overall sales revenue
* Monthly number of customers
* Monthly number of transactions per customer

### Load required libraries
```{r, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(tidyr)
```

### Read in data from previous Task
```{r}
filePath <- "C:/Users/Liz/Documents/Virtual_Internship/"
data <- fread(paste0(filePath,"QVI_data.csv"))
```

### Set themes for plots
```{r}
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
```

# Select control stores
## Calculate measures over time for each store
### Add a new month ID column in the data with the format yyyymm
```{r}
data[, YEARMONTH := format(DATE,'%Y%m')]
```

### Measure calculations for each store and month
```{r}
measureOverTime <- data[, .(totSales = sum(TOT_SALES),
                            nCustomers = uniqueN(LYLTY_CARD_NBR),
                            nTxnPerCust = .N/uniqueN(LYLTY_CARD_NBR),
                            nChipsPerTxn = sum(PROD_QTY)/uniqueN(LYLTY_CARD_NBR),
                            avgPricePerUnit = sum(TOT_SALES)/sum(PROD_QTY))
                        , .(YEARMONTH , STORE_NBR)][order(YEARMONTH)]
```

### Filter to the stores with full observation periods and the pre-trial period
```{r}
storesWithFullObs <- unique(measureOverTime[, .N, STORE_NBR][N == 12, STORE_NBR])
preTrialMeasures <- measureOverTime[YEARMONTH < 201902 & STORE_NBR %in% storesWithFullObs, ]
```

## Functions
### Function to calculate correlation
We need to work out a way of ranking how similar each potential control store is to the trial store. We can calculate how correlated the performance of each store is to the trial store.
```{r}
calculateCorrelation <- function(inputTable, metricCol, storeComparison) {
  
  calcCorrTable = data.table(Store1 = numeric(), Store2 = numeric(), corr_measure = numeric())
  storeNumbers <- unique( inputTable[, STORE_NBR ] )
  
    for (i in storeNumbers) {
      calculatedMeasure = data.table("Store1" = storeComparison,
                                     "Store2" = i,
                                     "corr_measure" = cor(inputTable[STORE_NBR == i, ..metricCol],
                                                          inputTable[STORE_NBR == storeComparison, ..metricCol])
      )
      calcCorrTable <- rbind(calcCorrTable, calculatedMeasure, use.names = FALSE)
    }
  return(calcCorrTable)
}
```

### Function to calculate a standardised magnitude distance for a measure
Apart from correlation, we can also calculate a standardised metric based on the absolute difference between the trial store's performance and each control store's performance.
```{r}
calculateMagnitudeDistance <- function(inputTable, metricCol, storeComparison) {
  
  calcDistTable = data.table(Store1 = numeric(), Store2 = numeric(), YEARMONTH = numeric(), measure = numeric())
  storeNumbers <- unique(inputTable[, STORE_NBR])
  
  for (i in storeNumbers) {
    calculatedMeasure = data.table("Store1" = storeComparison
                                   , "Store2" = i
                                   , "YEARMONTH" = inputTable[STORE_NBR == storeComparison, YEARMONTH]
                                   , "measure" = abs(inputTable[STORE_NBR == storeComparison, eval(metricCol)]
                                                     - inputTable[STORE_NBR == i, eval(metricCol)])
    )
    calcDistTable <- rbind(calcDistTable, calculatedMeasure)
  }
    # Standardise the magnitude distance so that the measure ranges from 0 to 1
  minMaxDist <- calcDistTable[, .(minDist = min(measure), maxDist = max(measure)), by = c("Store1", "YEARMONTH")]
  distTable <- merge(calcDistTable, minMaxDist, by = c("Store1", "YEARMONTH"))
  distTable[, magnitudeMeasure := 1 - (measure - minDist)/(maxDist - minDist)]
  finalDistTable <- distTable[, .(mag_measure = mean(magnitudeMeasure)), by =
                                .(Store1, Store2)]
  return(finalDistTable)
}
```

# Store 77

## Use the functions to find the control stores
We'll select control stores based on how similar monthly total sales in dollar amounts and monthly number of customers are to the trial stores. So we will need to use our functions to get four scores, two for each of total sales and total customers.

### Calculate correlations against store 77 using total sales and number of customers.
```{r}
trial_store <- 77
corr_nSales <- calculateCorrelation(preTrialMeasures, 'totSales', trial_store)
corr_nCustomers <- calculateCorrelation(preTrialMeasures, 'nCustomers', trial_store)
```

### Calculating magnitude.
```{r}
magnitude_nSales <- calculateMagnitudeDistance(preTrialMeasures, quote(totSales), trial_store)
magnitude_nCustomers <- calculateMagnitudeDistance(preTrialMeasures, quote(nCustomers), trial_store)
```

### Combine all the scores calculated to create a composite score to rank on.
```{r}
corr_weight <- 0.5
score_nSales <- merge(corr_nSales, magnitude_nSales, by = c('Store1','Store2'))[
                , scoreNSales := corr_weight*corr_measure + corr_weight*mag_measure]


score_nCustomers <- merge(corr_nCustomers, magnitude_nCustomers, by = c('Store1','Store2'))[
                , scoreNCust := corr_weight*corr_measure + corr_weight*mag_measure]
```
Now we have a score for each of total number of sales and number of customers.
Let's combine the two via a simple average.

### Combine scores across the drivers
```{r}
score_Control <- merge(score_nSales, score_nCustomers, by = c('Store1','Store2'))
score_Control[, finalControlScore := scoreNSales * 0.5 + scoreNCust * 0.5]
```

### Control store
The store with the highest score is then selected as the control store since it is most similar to the trial store.
```{r}
controlStores <- score_Control[finalControlScore > 0.85, .(Store2, finalControlScore)]
controlStores
control_store <- 233
```

Now that we have found a control store, let's check visually if the drivers are indeed similar in the period before the trial.

## Visual checks on trends based on the drivers before de trial
### Total sales
```{r}
measureOverTimeSales <- copy(measureOverTime)
pastSales <- measureOverTimeSales[, Store_type := ifelse(STORE_NBR == trial_store, "Trial",
                                                  ifelse(STORE_NBR == control_store, "Control", "Other stores"))][
  , totalSales := mean(totSales), by = c("YEARMONTH","Store_type")][
  , TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1, sep = "-"),"%Y-%m-%d")][
      as.numeric(YEARMONTH) < 201903 , ]

ggplot(pastSales, aes(TransactionMonth, totalSales, color = Store_type)) +
  geom_line() +
  labs(x = "Month of operation", y = "Total sales", title = "Total sales by month")
```

### Number of customers.
```{r}
measureOverTimeCusts <- copy(measureOverTime)
pastCustomers <- measureOverTimeCusts[,Store_type := ifelse(STORE_NBR == trial_store, "Trial",
                                                     ifelse(STORE_NBR == control_store, "Control", "Other stores"))][
  , totalCustomers := mean(nCustomers), by = c('YEARMONTH','Store_type')][
  , TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1, sep = "-"),"%Y-%m-%d")][
      as.numeric(YEARMONTH) < 201903 , ]
  
ggplot(pastCustomers, aes(TransactionMonth, totalCustomers, color = Store_type)) +
  geom_line() +
  labs(x = "Month of operation", y = "Number of customers", title = "Number of customers by month")

```

# Assessment of trial store 77
The trial period goes from the start of February 2019 to April 2019.

## Assessing for sales

We now want to see if there has been an uplift in overall chip sales.

### Scale pre-trial control sales to match pre-trial trial store sales
```{r}
scalingFactorForControlSales <- preTrialMeasures[STORE_NBR == trial_store &
                        YEARMONTH < 201902, sum(totSales)]/preTrialMeasures[STORE_NBR == control_store &
                        YEARMONTH < 201902, sum(totSales)]
```

### Apply the scaling factor
```{r}
scaledControlSales <- measureOverTimeSales[STORE_NBR == control_store, ][ ,
                    controlSales := totSales * scalingFactorForControlSales]
```

Now that we have comparable sales figures for the control store, we can calculate the percentage difference between the scaled control sales and the trial store's sales during the trial period.

### Calculate the percentage difference between scaled control sales and trial sales
```{r}
trialSales <- measureOverTimeSales[STORE_NBR == trial_store,]

percentageDiff <- merge(scaledControlSales[,c('YEARMONTH', 'controlSales')],
                        trialSales[,c('YEARMONTH','totSales')],
                        by = 'YEARMONTH')[, percentageDiff := (abs(controlSales-totSales)/
                                          ((controlSales+totSales)/2))]
```

Let's see if the difference is significant!
As our null hypothesis is that the trial period is the same as the pre-trial period, let's take the standard deviation based on the scaled percentage difference in the pre-trial period


```{r}
stdDev <- sd(percentageDiff[YEARMONTH < 201902 , percentageDiff])
# Note that there are 8 months in the pre-trial period hence 8 - 1 = 7 degrees of freedom
degreesOfFreedom <- 7
```

We will test with a null hypothesis of there being 0 difference between trial and control stores.
### Calculate the t-values for the trial months. 

```{r}
percentageDiff[, tValue := abs(controlSales-totSales)/stdDev][
    , TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1,
                                        sep = "-"),"%Y-%m-%d")]
percentil95 <-  qt(0.95, df = degreesOfFreedom)
```
The increase in sales in the trial store in March and April is statistically greater than in the control store.

Let's create a more visual version of this by plotting the sales of the control store, the sales of the trial stores and the 95th percentile value of sales of the control store.

### Trial and control store total sales
```{r}
pSales <- measureOverTimeSales[, Store_type := ifelse(STORE_NBR == trial_store, "Trial",
                                              ifelse(STORE_NBR == control_store, "Control", "Other stores"))][
    , totalSales := mean(totSales), by = c("YEARMONTH","Store_type")][
    , TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1, sep = "-"),"%Y-%m-%d")][
      Store_type %in% c("Trial", "Control"), ]
```

### Control store 95th percentile
```{r}
pSales_Controls95 <- pSales[Store_type == "Control",
][, totalSales := totalSales * (1 + stdDev * 2)
][, Store_type := "Control 95th % confidence
interval"]
```

### Control store 5th percentile
```{r}
pSales_Controls5 <- pSales[Store_type == "Control",
][, totalSales := totalSales * (1 - stdDev * 2)
][, Store_type := "Control 5th % confidence
interval"]
```

```{r}
trialAssessment <- rbind(pSales, pSales_Controls95, pSales_Controls5)
```

### Plotting these in one nice graph
```{r}
ggplot(trialAssessment, aes(TransactionMonth, totalSales, color = Store_type)) +
  geom_rect(data = trialAssessment[ YEARMONTH < 201905 & YEARMONTH > 201901 ,],
            aes(xmin = min(TransactionMonth), xmax = max(TransactionMonth), ymin = 0 , ymax =
                  Inf, color = NULL), show.legend = FALSE) +
  geom_line() +
  labs(x = "Month of operation", y = "Total sales", title = "Total sales by month")
```

Trial in store 77 is significantly different to its control store in the trial period in two of the three trial months.

## Assessing for number of customers
### Scaling Factor
```{r}
scalingFactorForControlCustomers <- preTrialMeasures[STORE_NBR == trial_store & YEARMONTH < 201902, sum(nCustomers)]/
                                    preTrialMeasures[STORE_NBR == control_store & YEARMONTH < 201902, sum(nCustomers)]
```

### Apply the scaling factor
```{r}
scaledControlCusts <- measureOverTimeCusts[STORE_NBR == control_store, 
                                           ][ ,controlCustomers := nCustomers * scalingFactorForControlCustomers]
```

### Calculate the percentage difference between scaled control customers and trial customers
```{r}
trialCusts <- measureOverTimeCusts[STORE_NBR == trial_store,]

percentageDiffCusts <- merge(scaledControlCusts[,c('YEARMONTH', 'controlCustomers')],
                        trialCusts[,c('YEARMONTH','totalCustomers')],
                        by = 'YEARMONTH')[, percentageDiff := (abs(controlCustomers-totalCustomers)/
                                                                 ((controlCustomers+totalCustomers)/2))]
```
Let's see if the difference is significant

```{r}
stdDevCust <- sd(percentageDiffCusts[YEARMONTH < 201902 , percentageDiff])
```

### Calculate the t-values for the trial months.
```{r}
percentageDiffCusts[, tValue := abs(controlCustomers-totalCustomers)/stdDevCust
                    ][, TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1,
                                      sep = "-"),"%Y-%m-%d")]

```

### Trial and control store number of customers
```{r}
pCusts <- measureOverTimeCusts[, Store_type := ifelse(STORE_NBR == trial_store, "Trial",
                                                      ifelse(STORE_NBR == control_store, "Control", "Other stores"))
       ][, totalCustomers2 := mean(totalCustomers), by = c("YEARMONTH","Store_type")
       ][, TransactionMonth := as.Date(paste(as.numeric(YEARMONTH)%/%100, as.numeric(YEARMONTH)%%100, 1, sep = "-"),"%Y-%m-%d")
       ][Store_type %in% c("Trial", "Control"), ]
```
### Control store 95th percentile
```{r}
pCusts_Controls95 <- pCusts[Store_type == "Control",
][, totalCustomers2 := totalCustomers2 * (1 + stdDevCust * 2)
][, Store_type := "Control 95th % confidence
interval"]
```

### Control store 5th percentile
```{r}
pCusts_Controls5 <- pCusts[Store_type == "Control",
][, totalCustomers2 := totalCustomers2 * (1 - stdDevCust * 2)
][, Store_type := "Control 5th % confidence
interval"]

trialAssessmentCusts <- rbind(pCusts, pCusts_Controls95, pCusts_Controls5)
```

### Plotting these in one nice graph
```{r}
ggplot(trialAssessmentCusts, aes(TransactionMonth, totalCustomers2, color = Store_type)) +
  geom_rect(data = trialAssessmentCusts[ YEARMONTH < 201905 & YEARMONTH > 201901 ,],
            aes(xmin = min(TransactionMonth), xmax = max(TransactionMonth), ymin = 0 , ymax =
                  Inf, color = NULL), show.legend = FALSE) +
  geom_line() +
  labs(x = "Month of operation", y = "Total customers", title = "Total customers by month")
```
